---
title: SPOROS pipeline analysis of seed composition data using multinomial mixed models
output: 
  html_document:
    number_sections: TRUE
graphics: yes
---

```{r setup, include=FALSE}
library("knitr")
library(SASmarkdown)
library("ggplot2")
library("ggnewscale")
library("tidyverse")
library("readxl")

options(round = 4)
options(stringsAsFactors = FALSE)

opts_chunk$set(echo = TRUE, include=TRUE,
               comment = "  ", prompt=TRUE,  
               warning = FALSE, message = FALSE,
               fig.width = 8.5, fig.asp=.8) 

# set path where the data for SAS will be saved to 
datapath <- "G:\\PeterM_XXX\\Analysis\\Data\\MethodsPaper\\"
plotpath <- "G:\\PeterM_XXX\\Analysis\\Code\\MethodsPaper\\"

```


# Read in Output F from GitHub

```{r}
## read in raw data from GitHub
f.dicer <- "https://raw.githubusercontent.com/ebartom/SPOROS/main/Figure2.SPOROSpaper/totalCounts/sRNA/F_seedExpand.DicerKO.avg.sRNA.Figure2.txt"
f.drosha <- "https://raw.githubusercontent.com/ebartom/SPOROS/main/Figure2.SPOROSpaper/totalCounts/sRNA/F_seedExpand.DroshaKO.avg.sRNA.Figure2.txt"
f.wildtype <- "https://raw.githubusercontent.com/ebartom/SPOROS/main/Figure2.SPOROSpaper/totalCounts/sRNA/F_seedExpand.Wildtype.avg.sRNA.Figure2.txt"

dicer <- read.table(f.dicer, header=TRUE, sep="\t")
drosha <- read.table(f.drosha, header=TRUE, sep="\t")
wildtype <- read.table(f.wildtype, header=TRUE, sep="\t")
```



```{r include=FALSE}
# add sample name to each dataset
dicer$Sample <- "DicerKO"
drosha$Sample <- "DroshaKO"
wildtype$Sample <- "Wildtype"

# if needed, update SeedID to make it unique across samples
if(any(dicer$SeedID %in% drosha$SeedID) | 
   any(dicer$SeedID %in% wildtype$SeedID) |
   any(drosha$SeedID %in% wildtype$SeedID)){
  dicer$SeedID <- paste("DicerKO", dicer$SeedID, sep=".")
  drosha$SeedID <- paste("DroshaKO", drosha$SeedID, sep=".")
  wildtype$SeedID <- paste("Wildtype", wildtype$SeedID, sep=".")
}

# combine data
dataCombined <- rbind(dicer, drosha, wildtype) %>% as.data.frame


```

There are `r (dicer %>% nrow)/6` seeds in DicerKO data.

There are `r (drosha %>% nrow)/6` seeds in DroshaKO data.

There are `r (wildtype %>% nrow)/6` seeds in Wildtype data.

There are `r length(unique(dataCombined$SeedID))` seeds in the combined data set.


Example of each data set:
```{r }
# look at each file to check that data were read and combined correctly
dicer %>% head
drosha %>% head
wildtype %>% head
dataCombined %>% head
dataCombined %>% tail

```

Do the counts match? 
```{r}
# check total seeds/id's
(nrow(dicer)/6 + nrow(drosha)/6 + nrow(wildtype)/6 ==
   length(unique(dataCombined$SeedID)))

```



# Save the data

```{r echo=FALSE, include=FALSE}
### save data for SAS use
fileOut <- paste(paste(datapath, "data.seedLong", sep=""),
                 Sys.Date(), "csv", sep=".")

write.table(x = dataCombined,
            file = fileOut,
            sep = ",",
            quote = FALSE,
            row.names = FALSE)
```


Data saved for SAS analyses into:
```{r echo=FALSE}
fileOut
```

# SAS Analysis using PROC GLIMMIX

```{r include=FALSE}
# https://www.r-bloggers.com/2019/09/create-rmarkdown-document-with-sas-code-and-output-sas-engine/

saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -nocenter"
knitr::opts_chunk$set(
  comment=NA,
  echo=TRUE)
```


```{r, engine="sashtml", engine.path=saspath, engine.opts=sasopts, collectcode=TRUE}
* update the path in the INFILE statement to match datapath above;
* note that sas uses single backslash in paths, e.g. "C:\sasdata";

data seedlong;
informat seed $6. sample $20. seedid $30.;
infile "G:\\PeterM_XXX\\Analysis\\Data\\MethodsPaper\\data.seedLong.2021-08-12.csv" 
  dlm="," firstobs=2;
input Seed $ sample $ seedid $ pos base $ ;
run;

proc glimmix data=seedlong outdesign = xx method=rmpl;
class seed seedid sample(ref="Wildtype") pos base(ref="A");
model base = sample pos sample*pos /dist=multinomial link=glogit s or(label) ddfm=bw;
random intercept/ subject = seedid group=base type=chol;
store gmxres;
run;
```

```{r, engine="sashtml", engine.path=saspath, engine.opts=sasopts, collectcode=TRUE, results="hide"}

proc plm restore=gmxres noclprint plots=none;
lsmeans sample/ilink oddsratio adj=tukey cl e;
slice sample*pos/sliceby=pos diff oddsratio adj=tukey cl;
ods output slicediffs= sampleposdiffs diffs= samplediffs;
run;

```

```{r, engine="sashtml", engine.path=saspath, engine.opts=sasopts, collectcode=TRUE}
proc sort data=sampleposdiffs;
by sample slice base;

proc print data=sampleposdiffs;
where _sample = "Wildtype";
var sample _sample slice base OddsRatio AdjLowerOR AdjUpperOR Adjp;
run;      
```

```{r, engine="sashtml", engine.path=saspath, engine.opts=sasopts, collectcode=TRUE}
* update the path in the OUTFILE statement below to match datapath;
* note that sas uses single backslash for paths, e.g. "C:\sasdata";

proc export data=work.sampleposdiffs
  outfile="G:\PeterM_XXX\Analysis\Data\MethodsPaper\F_OREstimates_Fig2_081221.xlsx"
  dbms=excel replace;
  sheet="OR Position";
run;
```




```{r include=FALSE}
# read in the OR estimates data file that was output by PROC EXPORT above
orest <- read_excel(paste(datapath, "F_OREstimates_Fig2_081221.xlsx", sep=""))

# change options to allow better pvalue formatting for small p
scipen0 <- getOption("scipen")
options(scipen = 10000000)

# add additional columns used in plotting
orest <- orest %>% 
  filter(`_sample` == "Wildtype") %>% 
  mutate(pos = Slice,
         pos2 = substr(pos, 5,5),
         OR = ifelse(OddsRatio>1, OddsRatio, 1/OddsRatio),
         iUp = ifelse(OddsRatio > 1, ">A", "<A"),
         up = ifelse(OddsRatio > 1, base, "A"),
         base2 = factor(base), 
         base2 = factor(base2, levels = rev(levels(base2))),
         orlbl = paste(format(OR, digits=2),
                        "\n(", format.pval(Adjp, eps=0.0001, digits=2), ")", 
                        sep=""),
         # boldpval = ifelse(Adjp < .05, "bold", "plain"),
         psignif = ifelse(Adjp < .05, "signif", "notsignif"))

# reset significant digits display option
options(scipen = scipen0)

# create separate data.frames for plotting
dicerKO <- orest %>% 
  filter(sample == "DicerKO") 

droshaKO <- orest %>% 
  filter(sample == "DroshaKO") 


```


# Dicer vs. Wildtype


```{r, echo=FALSE, fig.width=11, fig.height=8.5}
maxsize <- dicerKO$OR %>% max
ggplot(data=dicerKO, aes(pos2, base2)) + 
  geom_point(aes(size = OR, color = up)) +
  scale_color_manual(values = c("A" = "#0100CC",
                                "C" = "#FFB302",
                                "G" = "#CC0000",
                                "U" = "#01CC00"), 
                     name=" ",
                     guide = guide_legend(override.aes = list(size = 5))) +
  new_scale_color() +
  geom_point(aes(size = maxsize, color = psignif), shape=21, stroke=2) +
  scale_color_manual(values = c("signif" = "black",
                                "notsignif" = "gray"), 
                     guide =  FALSE) +
  scale_size(range = c(0, 50), # update upper limit to change circle size
             breaks = c(2, 5, 15, 25),
             name="OR") +
  theme_bw(base_size = 20) +
  ylab("Base vs. A") +
  xlab("Position") +
  ggtitle("Dicer vs. Wildtype") + 
  guides(size = guide_legend(order = 1, 
                             override.aes=list(colour = "gray", shape = 19))) + 
  theme(legend.box.just = "center", legend.position = "bottom",
        legend.box.margin = margin(t = -14))

# note: scaling may appear different in the PDF file due to specified height/width
ggsave(paste(plotpath, "F_Dicer_OddsRatioPlot_081221.pdf", sep=""),
       width=11, height=8.5, units="in")

# notes:
# scale_color_manual() creates a legend for A, C, G, U color to match weblogos
# new_scale_color() in ggnewscale package resets scale_clor_manual and allows multiple
#     custom new_scale_color() statements
# scale_size() scales circles so that area is proportional to specified value
# scales and font sizes are optimized for 8.5x11 inch PDF file viewing

```

# Drosha vs. Wildtype



```{r, echo=FALSE, fig.width=11, fig.height=8.5}
maxsize <- droshaKO$OR %>% max
ggplot(data=droshaKO, aes(pos2, base2)) + 
  geom_point(aes(size = OR, color = up)) +
  scale_color_manual(values = c("A" = "#0100CC",
                                "C" = "#FFB302",
                                "G" = "#CC0000",
                                "U" = "#01CC00"), 
                     name=" ",
                     guide = guide_legend(override.aes = list(size = 5))) +
  new_scale_color() +
  geom_point(aes(size = maxsize, color = psignif), shape=21, stroke=2) +
  scale_color_manual(values = c("signif" = "black",
                                "notsignif" = "gray"), 
                     guide =  FALSE) +
  scale_size(range = c(0, 50), # update upper limit to change circle size
             breaks = c(2, 5, 15, 30),
             name="OR") +
  theme_bw(base_size = 20) +
  ylab("Base vs. A") +
  xlab("Position") +
  ggtitle("Drosha vs. Wildtype") + 
  guides(size = guide_legend(order=1, 
                             override.aes=list(colour = "gray", shape = 19))) + 
  theme(legend.box.just = "center", legend.position = "bottom",
        legend.box.margin = margin(t = -14))

# note: scaling may appear different in the PDF file due to specified height/width
ggsave(paste(plotpath, "F_Drosha_OddsRatioPlot_081221.pdf", sep=""),
       width=11, height=8.5, units="in")


```
